# Linear Regression

Another task of the project is to perform a regression, to predict _median home value_ based on a suitable predictor.

It has been shown that the target variable exhibits a correlation with the _average rooms_ one. Since they seem to align in a linear fashion, a regression of linear would suit the needs.

```{r}
# linear regression
df.housing.imputed.refined |>
  ggplot(aes(average_rooms, median_home_value)) +
  geom_point(
    shape = 21,
    size = 2.5,
    color = "cornflowerblue"
  ) +
  geom_smooth(
    formula = y ~ x,
    method = "lm",
    se = FALSE,
    col = "coral",
    fullrange = TRUE
  ) +
  scale_x_continuous(
    limits = c(0, 12),
    breaks = 0:12
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 5)
  ) +
  labs(
    title = "Linear regression scatterplot of Average Rooms vs Median Home Value",
    x = "Average Rooms",
    y = "Median Home Value"
  )
```

```{r}
# linear regression model

mdl.linear_regression <- lm(
  formula = median_home_value ~ average_rooms,
  data = df.housing.imputed.refined
)

mdl.linear_regression |>
  summary()
```

According to the linear regression model, the regression line intercepts the ordinate at $3.7854$ and has a slope of $7.1886$, from which the following regression equation can be drawn:
  
$$
\begin{split}
\hat{y}(x) = \beta_0 + \beta_1 x \\
\hat{y}(x)  = 3.7854 + 7.1886 x
\end{split}
$$

With missing values being replaced, the coefficient of determination ($R^2$) indicates that around $75\%$ of the variation of the dependent variable _(median home value)_ can be accounted for the independent variable _(average rooms)_.

This linear regression model indicates that, on average, each additional _room_ results in a $\$7188.6$ increase in the _median home value_.
